---
title: "Exercise - Block 7: Finalizing Figures & Reproducibility"
format: html
editor: source
editor_options:
  chunk_output_type: console
---

## Learning Objectives

Upon completing this exercise block, you will be able to:

-   **Produce and export publication-quality figures:**
    -   **Adapt** stylistic elements like fonts and line weights to meet publication guidelines.
    -   **Generate** high-resolution figures in various formats (PDF, PNG, TIFF) using functions like `ggsave()`.
-   **Implement best practices for reproducibility:**
    -   **Write** clean and reusable R scripts for figure generation.
    -   **Document** code and workflows to ensure reproducibility and transparency.

## Introduction

In this block, we will focus on finalizing figures for publication and ensuring reproducibility. We will cover how to adapt stylistic elements, generate high-resolution figures, and implement best practices for writing clean and reusable R scripts.

---

## Exercise 1: Exporting Figures

In this exercise, we will practice exporting the figures we created in the previous blocks.

**A.** Go back to `exercise_block_6.qmd`. Export the combined plot from "Exercise 4: Advanced setting of height and width of indivisual plots and combining all plots" as a high-resolution PNG file.

**B.** Export the same figure as a PDF file.

::: {.callout-tip collapse="true"}

### Solution
```{r, eval=FALSE}
# Add this code to the end of the solution for Exercise 4 in exercise_block_6.qmd

# Export as PNG
png("combined_figure.png", width = 10, height = 8, units = "in", res = 300)
gridExtra::grid.arrange(p1, ht_grob, img_draw, mm_res, ncol = 2)
dev.off()

# Export as PDF
pdf("combined_figure.pdf", width = 10, height = 8)
gridExtra::grid.arrange(p1, ht_grob, img_draw, mm_res, ncol = 2)
dev.off()
```
:::

---

## Exercise 2: Reproducible Scripting

In this exercise, we will focus on making our R scripts more reproducible.

**A.** Review the R code in `exercise_block_6.qmd`. Identify any areas where comments could be improved or added to enhance clarity and reproducibility.

**B.** Suggest ways to make the data loading and processing steps more robust (e.g., error handling, clearer variable names).

**C.** Create a function that encapsulates the creation of the complex heatmap from `exercise_block_6.qmd`. The function should take the data as input and return the heatmap object.

::: {.callout-tip collapse="true"}

### Solution
**A. Improved Comments:**
- Ensure every major step (data loading, scaling, annotation creation, plotting) has a clear, concise comment explaining its purpose.
- Add comments to complex lines of code to explain the logic.

**B. More Robust Scripting:**
- **Error Handling:** Use `tryCatch()` for data loading to gracefully handle cases where the URL might be inaccessible.
- **Clearer Variable Names:** While `mat` and `lfc_data` are common in R, consider more descriptive names like `expression_matrix_scaled` or `log_fold_change_data`.
- **Function Encapsulation:** For complex plotting routines, consider wrapping them in functions to improve reusability and readability.
- **Configuration:** If parameters like the number of top genes or LFC thresholds change frequently, consider defining them as variables at the top of the script.

**C. Function for Complex Heatmap:**
```{r, eval=FALSE}
create_complex_heatmap <- function(data) {
  # Prepare the data
  mat <- as.matrix(data$counts)
  mat_scaled <- t(scale(t(mat))) # Z-score scaling
  gene_vars <- apply(mat_scaled, 1, var, na.rm = TRUE)
  top_genes_var <- names(sort(gene_vars, decreasing = TRUE)[1:50])
  mat_top_var <- mat_scaled[top_genes_var, ]

  # Create a column annotation with custom colors
  col_anno <- HeatmapAnnotation(
    compound = data$anno$compound,
    time = data$anno$time,
    col = list(
      compound = c("none" = "steelblue", "sucrose" = "tomato"),
      time = c("3 hour" = "lightgrey", "24 hour" = "darkgrey")
    )
  )

  # Extract genes matching the profile
  profile_genes <- data$diff %>%
    filter(
      sucrose_3h_lfc > 1 & sucrose_3h_pval < 0.05 |
        sucrose_24h_lfc < -1 & sucrose_24h_pval < 0.05
    ) %>%
    pull(Gene_ID)

  mat_profile <- mat_scaled[profile_genes, ]

  # Create a row annotation
  rownames(data$diff) <- data$diff$Gene_ID
  lfc_data <- data$diff[profile_genes, c("sucrose_3h_lfc", "sucrose_24h_lfc")]

  row_anno <- rowAnnotation(
    "3h LFC" = anno_barplot(
      lfc_data[, 1],
      gp = gpar(fill = "tomato"),
      axis_param = list(at = c(-2, 0, 2))
    ),
    "24h LFC" = anno_barplot(
      lfc_data[, 2],
      gp = gpar(fill = "steelblue"),
      axis_param = list(at = c(-2, 0, 2))
    ),
    width = unit(4, "cm")
  )

  # Combine with the heatmap
  ht <- Heatmap(mat_profile,
    name = "Z-Score",
    top_annotation = col_anno,
    show_row_names = FALSE,
    column_title = "Genes Up at 3h, Down at 24h"
  ) + row_anno
  
  return(ht)
}

# Example usage:
# ht_object <- create_complex_heatmap(data)
```
:::

## Session information
::: {.callout-tip collapse="true"}
```{r}
sessionInfo()
```
:::
