---
title: "Exercise - Block 7: Capstone Project"
format: html
editor: source
editor_options:
  chunk_output_type: console
---

## Learning Objectives

Upon completing this capstone exercise, you will be able to:

-   **Synthesize** skills from across the course to create a complex, multi-panel, publication-ready figure.
-   **Combine** plots from `ComplexHeatmap` and `ggplot2` into a cohesive layout using `grid` graphics.
-   **Apply** consistent styling and themes across different plot types.
-   **Export** final figures in both high-resolution raster (TIFF) and vector (PDF) formats, adhering to typical journal specifications.

## Introduction

In this final block, we will undertake a capstone project that brings together everything we have learned. We will create a publication-ready, multi-panel figure from scratch. This figure will include a complex heatmap, a volcano plot, and a boxplot, all arranged in a clean layout and styled consistently.

---

## Packages and Data

First, let's load the necessary packages and the course dataset.

```{r, message=FALSE, warning=FALSE}
# Load packages
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)
library(grid)
library(RColorBrewer)

# Load data
data <- readRDS(gzcon(url(
  "https://raw.githubusercontent.com/urppeia/publication_figs/main/data.rds"
)))

rownames(data$counts) <- data$counts$Gene_ID
data$counts <- data$counts[, -1]
```

---

## Capstone Exercise: Creating a Publication-Ready Figure

### Part A: Top Panel - Complex Heatmap

First, we will create a heatmap of the top 50 most variable genes, annotated by experimental conditions.

::: {.callout-tip collapse="true"}

### Solution

```{r}
# 1. Prepare the data: scale and find top 50 variable genes
mat <- as.matrix(data$counts)
mat_scaled <- t(scale(t(mat)))
gene_vars <- apply(mat_scaled, 1, var, na.rm = TRUE)
top_genes_var <- names(sort(gene_vars, decreasing = TRUE)[1:50])
mat_top_var <- mat_scaled[top_genes_var, ]

# 2. Create column annotation
col_anno <- HeatmapAnnotation(
  compound = data$anno$compound,
  time = data$anno$time,
  col = list(
    compound = c("none" = "steelblue", "sucrose" = "tomato"),
    time = c("3 hour" = "lightgrey", "24 hour" = "darkgrey")
  )
)

# 3. Create the heatmap object
ht <- Heatmap(mat_top_var,
  name = "Z-Score",
  top_annotation = col_anno,
  show_row_names = FALSE,
  column_title = "Top 50 Most Variable Genes"
)

# 4. Draw it to check
draw(ht)
```
:::

### Part B: Bottom-Left Panel - Volcano Plot

Next, we'll create a volcano plot for the 24-hour differential expression data to visualize up- and down-regulated genes.

::: {.callout-tip collapse="true"}

### Solution

```{r}
# 1. Add a significance column to the differential expression data
data$diff <- data$diff %>%
  mutate(significant = sucrose_24h_pval < 0.05 & abs(sucrose_24h_lfc) > 1)

# 2. Create the volcano plot
p_volcano <- ggplot(data$diff, aes(x = sucrose_24h_lfc, y = -log10(sucrose_24h_pval))) +
  geom_point(aes(color = significant), alpha = 0.6) +
  scale_color_manual(values = c("grey", "red")) +
  theme_bw() +
  labs(
    title = "Volcano Plot (24h)",
    x = "Log2 Fold Change",
    y = "-log10(p-value)"
  ) +
  theme(legend.position = "none")

# 3. Print it to check
print(p_volcano)
```
:::

### Part C: Bottom-Right Panel - Boxplot

Now, let's focus on a single, highly significant gene and show its expression across samples with a boxplot. We'll pick the gene with the lowest p-value at 24 hours.

::: {.callout-tip collapse="true"}

### Solution

```{r}
# 1. Find the top significant gene at 24h
top_gene_24h <- data$diff %>%
  filter(sucrose_24h_pval < 0.05) %>%
  arrange(sucrose_24h_pval) %>%
  pull(Gene_ID) %>%
  head(1)

# 2. Prepare data for plotting
plot_data <- data.frame(
  expression = as.numeric(mat[top_gene_24h, ]),
  compound = data$anno$compound,
  time = data$anno$time
)

# 3. Create the boxplot
p_boxplot <- ggplot(plot_data, aes(x = compound, y = expression, fill = time)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = paste("Expression of", top_gene_24h),
    x = "Compound",
    y = "Normalized Expression"
  ) +
  scale_fill_manual(values = c("lightgrey", "darkgrey"))

# 4. Print it to check
print(p_boxplot)
```
:::

### Part D: Assembling the Final Figure

With all three components ready, we will now use the `grid` package to assemble them into a single, well-proportioned figure. The top panel will contain the heatmap, and the bottom panel will be split between the volcano plot and the boxplot.

::: {.callout-tip collapse="true"}

### Solution

```{r, fig.height=8, fig.width=8}
# 1. Grab the heatmap as a graphical object (grob)
ht_grob <- grid.grabExpr(draw(ht))

# 2. Clear the graphics device and set up the layout
grid.newpage()
# Layout: 2 rows, 2 columns. Heatmap spans both columns in row 1.
pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 2, heights = unit(c(0.5, 0.5), "null"))))

# 3. Place the heatmap in the top row (spanning columns 1 and 2)
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1:2))
grid.draw(ht_grob)
popViewport()

# 4. Place the volcano plot in the bottom-left
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))
print(p_volcano, newpage = FALSE)
popViewport()

# 5. Place the boxplot in the bottom-right
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2))
print(p_boxplot, newpage = FALSE)
popViewport()

popViewport()
```
:::

### Part E: Exporting for Publication

Finally, we will save our composite figure in two standard publication formats: a high-resolution TIFF and a vector-based PDF.

::: {.callout-tip collapse="true"}

### Solution

```{r}
# 1. Save as a 300 DPI TIFF file
tiff("capstone_figure.tiff", width = 8, height = 8, units = "in", res = 300)

# Re-draw the plot for the file device
grid.newpage()
pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 2, heights = unit(c(0.5, 0.5), "null"))))
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1:2))
grid.draw(ht_grob)
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))
print(p_volcano, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2))
print(p_boxplot, newpage = FALSE)
popViewport()
popViewport()

# Close the device
dev.off()


# 2. Save as a PDF file
pdf("capstone_figure.pdf", width = 8, height = 8)

# Re-draw the plot for the file device
grid.newpage()
pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 2, heights = unit(c(0.5, 0.5), "null"))))
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1:2))
grid.draw(ht_grob)
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))
print(p_volcano, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2))
print(p_boxplot, newpage = FALSE)
popViewport()
popViewport()

# Close the device
dev.off()
```
:::

## Session information
::: {.callout-tip collapse="true"}
```{r}
sessionInfo()
```
:::