---
title: "Exercise - Block 3"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

## Introduction to ggplot2 advanced topics

In the last block, we learned the basics of `ggplot2`. In this block, we will explore more advanced topics to make our plots publication-ready. We will cover:

*   **Faceting**: Creating subplots to compare different subsets of your data.
*   **Themes**: Customizing the appearance of your plots.
*   **Colors**: Using colors effectively and exploring different color palettes.
*   **Interactivity**: Making your plots interactive with `ggplotly`.

## Installing packages

```{r, eval=FALSE}
renv::install("RColorBrewer")
renv::install("viridis")
renv::install("dplyr")
renv::install("plotly")
```

## Packages
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(viridis)
library(dplyr)
library(plotly)
```

## Exploratory data analysis

```{r}
data <- readRDS(gzcon(url(
  "https://raw.githubusercontent.com/urppeia/publication_figs/main/data.rds"
)))
```

## Faceting

In Block 2, we created two separate volcano plots for the 3h and 24h timepoints. We can use faceting to create a single plot with two panels.

### Exercise 1

Create a single volcano plot faceted by timepoint. Store the plot as an object named `p1`.

::: {.callout-tip collapse="true"}
```{r}
# Create a single data frame for both timepoints
diff_df <- data.frame(
  lfc = c(data$diff$sucrose_3h_lfc, data$diff$sucrose_24h_lfc),
  pval = c(data$diff$sucrose_3h_pval, data$diff$sucrose_24h_pval),
  timepoint = rep(c("3h", "24h"), each = nrow(data$diff))
)
diff_df$significant <- diff_df$pval < 0.05 & abs(diff_df$lfc) > 1

p1 <- ggplot(diff_df, aes(x = lfc, y = -log10(pval), color = significant)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c("gray70", "tomato")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  labs(
    title = "Volcano Plot — Sucrose Treatment",
    x = "log2 Fold Change",
    y = "-log10(p-value)"
  ) +
  facet_wrap(~timepoint)

p1
```
:::

## Themes

`ggplot2` comes with several built-in themes. You can also create your own themes to have full control over the appearance of your plots.

### Exercise 2

**A.** Apply the `theme_minimal()` to the faceted volcano plot. Store the plot as `p2a`.

::: {.callout-tip collapse="true"}
```{r}
p2a <- p1 + theme_minimal()
p2a
```
:::

**B.** Customize the theme. Make the following changes:
*   Increase the title font size to 16.
*   Remove the panel grid lines.
*   Change the panel background to light gray.
Store the plot as `p2b`.

::: {.callout-tip collapse="true"}
```{r}
p2b <- p2a +
  theme(
    plot.title = element_text(size = 16),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "lightgray", color = NA)
  )
p2b
```
:::

## Colors

Colors are a powerful tool in data visualization. `ggplot2` provides many ways to work with colors.

### Exercise 3

Let's go back to the boxplot of expression data.

```{r}
data_long <- merge(data$anno, reshape2::melt(data$counts))
```

**A.** Use a color palette from `RColorBrewer`. Store the plot as `p3a`.

::: {.callout-tip collapse="true"}
```{r}
p3a <- ggplot(data_long, aes(x = Sample_ID, y = value, fill = compound)) +
  geom_boxplot(outlier.shape = NA) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = "Normalized expression per sample",
    x = "Sample",
    y = "Expression"
  ) +
  scale_fill_brewer(palette = "Set2")
p3a
```
:::

**B.** Use the `viridis` color palette. Store the plot as `p3b`.

::: {.callout-tip collapse="true"}
```{r}
p3b <- ggplot(data_long, aes(x = Sample_ID, y = value, fill = compound)) +
  geom_boxplot(outlier.shape = NA) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = "Normalized expression per sample",
    x = "Sample",
    y = "Expression"
  ) +
  scale_fill_viridis_d()
p3b
```
:::

**C.** Specify colors manually. Store the plot as `p3c`.

::: {.callout-tip collapse="true"}
```{r}
p3c <- ggplot(data_long, aes(x = Sample_ID, y = value, fill = compound)) +
  geom_boxplot(outlier.shape = NA) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = "Normalized expression per sample",
    x = "Sample",
    y = "Expression"
  ) +
  scale_fill_manual(values = c("none" = "#66c2a5", "sucrose" = "#fc8d62"))
p3c
```
:::

## Saving plots for publication

When preparing figures for publication, journals often have strict requirements for file format, dimensions, and resolution.

### Exercise 4

Let's save the faceted volcano plot from Exercise 2B (`p2b`).

**A.** Save the plot as a PNG file with a width of 8 inches, a height of 4 inches, and a resolution of 300 dpi.

::: {.callout-tip collapse="true"}
```{r}
ggsave("volcano_plot.png", plot = p2b, width = 8, height = 4, dpi = 300)
```
:::

**B.** Save the plot as a PDF file. PDF is a vector format, which is ideal for publications as it can be scaled without losing quality.

::: {.callout-tip collapse="true"}
```{r}
ggsave("volcano_plot.pdf", plot = p2b, width = 8, height = 4)
```
:::

**C.** Save the plot as a TIFF file.

::: {.callout-tip collapse="true"}
```{r}
ggsave("volcano_plot.tiff", plot = p2b, width = 8, height = 4, dpi = 300)
```
:::

## Fine-tuning stylistic elements

Journals often have specific guidelines for fonts and line weights.

### Exercise 5

Let's modify the volcano plot to meet some hypothetical publication guidelines.

**A.** Change the font family to "Arial". Store the plot as `p5a`.

::: {.callout-tip collapse="true"}
```{r}
p5a <- p2b + 
  theme(text = element_text(family = "Arial"))
p5a
```
:::

**B.** Adjust the line weights. Make the axis lines thicker (size = 1) and the facet borders thinner (size = 0.5). Store the plot as `p5b`.

::: {.callout-tip collapse="true"}
```{r}
p5b <- p5a + 
  theme(
    axis.line = element_line(linewidth = 1),
    strip.background = element_rect(linewidth = 0.5)
  )
p5b
```
:::


## Additional Plot Types

### Exercise 6: Scatter plot with regression line

Let's explore the relationship between the expression of two samples.

**A.** Create a scatter plot of the expression of the first two samples in the `data$counts` dataframe. Store the plot as `p6a`.

::: {.callout-tip collapse="true"}
```{r}
# Create a dataframe with the first two samples
sample_df <- data.frame(
  sample1 = data$counts[, 2],
  sample2 = data$counts[, 3]
)

p6a <- ggplot(sample_df, aes(x = sample1, y = sample2)) +
  geom_point() +
  labs(
    title = "Sample Expression Correlation",
    x = "Sample 1 Expression",
    y = "Sample 2 Expression"
  )
p6a
```
:::

**B.** Add a linear regression line to the scatter plot. Store the plot as `p6b`.

::: {.callout-tip collapse="true"}
```{r}
p6b <- p6a +
  geom_smooth(method = "lm")
p6b
```
:::

### Exercise 7: Bar plot with error bars

Let's visualize the average expression of each compound.

**A.** Calculate the mean and standard deviation of expression for each compound.

::: {.callout-tip collapse="true"}
```{r}
summary_df <- data_long %>%
  group_by(compound) %>%
  summarise(
    mean_expr = mean(value),
    sd_expr = sd(value)
  )

print(summary_df)
```
:::

**B.** Create a bar plot of the mean expression with error bars representing the standard deviation. Store the plot as `p7`.

::: {.callout-tip collapse="true"}
```{r}
p7 <- ggplot(summary_df, aes(x = compound, y = mean_expr, fill = compound)) +
  geom_bar(stat = "identity") +
  geom_errorbar(
    aes(ymin = mean_expr - sd_expr, ymax = mean_expr + sd_expr),
    width = 0.2
  ) +
  labs(
    title = "Average Gene Expression by Compound",
    x = "Compound",
    y = "Mean Expression"
  ) +
  theme_minimal()
p7
```
:::

## Interactive plots with ggplotly

`ggplotly` is a function from the `plotly` package that converts a `ggplot` object into an interactive `plotly` object.

### Example

Let's take the volcano plot from Exercise 1 (`p1`) and make it interactive.

```{r}
ggplotly(p1)
```

Now, let's try it on the scatter plot from Exercise 6b (`p6b`).

```{r}
ggplotly(p6b)
```

### Exercise 8: Your turn!

**A.** Create a scatter plot of `lfc` vs `-log10(pval)` from the `diff_df` dataframe, colored by `timepoint`. Store it as an object called `p8`.

::: {.callout-tip collapse="true"}
```{r}
p8 <- ggplot(diff_df, aes(x = lfc, y = -log10(pval), color = timepoint)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Volcano Plot by Timepoint",
    x = "log2 Fold Change",
    y = "-log10(p-value)"
  )
p8
```
:::

**B.** Now, use `ggplotly` to make the plot interactive.

::: {.callout-tip collapse="true"}
```{r}
ggplotly(p8)
```
:::


**C.** What if one just want to see 24h dots?.

::: {.callout-tip collapse="true"}
One can simply click on 3h dot in the legend, and they will disappear.
:::

## Session information
::: {.callout-tip collapse="true"}
```{r}
sessionInfo()
```
:::