---
title: "Exercise - Block 6"
editor: source
editor_options: 
  chunk_output_type: console
---

## Learning Objectives

Upon completing this exercise block, you will be able to:

- **Create and combine figures from different graphics systems:**
  - **Construct** complex, annotated heatmaps using the `ComplexHeatmap` package.
  - **Combine** plots from different systems (e.g., `ComplexHeatmap` and `ggplot2`) into a cohesive figure using `grid` graphics.

## Introduction

In this block, we will explore one of the most common visualizations in genomics: the heatmap. We will use the `ComplexHeatmap` package, a powerful and flexible tool for creating publication-quality heatmaps. We will also learn how to combine a `ComplexHeatmap` with a `ggplot` plot into a single figure, a common task in data visualization.

## Installing packages

```{r, eval=FALSE}
renv::install("ComplexHeatmap")
renv::install("circlize")
renv::install("dplyr")
renv::install("ggplot2")
renv::install("grid")
renv::install("gganatogram")
renv::install("cowplot")
renv::install("patchwork")
renv::install("RColorBrewer")
renv::install("egg")
```

## Packages
```{r, message=FALSE, warning=FALSE}
library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(ggplot2)
library(grid)
library(gganatogram)
library(cowplot)
library(patchwork)
library(RColorBrewer)
library(egg)
```

## Exploratory data analysis

```{r}
data <- readRDS(gzcon(url(
  "https://raw.githubusercontent.com/urppeia/publication_figs/main/data.rds"
)))

rownames(data$counts) <- data$counts$Gene_ID
data$counts <- data$counts[, -1]
```

---

## Exercise 1: Constructing a Complex Heatmap

In this exercise, we will create a complex heatmap with multiple annotations. This will allow us to visualize the relationships between different variables in our data.

**A.** First, let's prepare the data. We will scale the data and select the top 50 most variable genes.

**B.** Next, we will create a `HeatmapAnnotation` object for the columns that shows the `compound` and `time` for each sample. We will also define custom colors for the annotation levels.

**C.** Now, let's "precision extract" a gene profile. We will find and visualize genes that follow a very specific biological pattern: genes that are activated early and then repressed later.

**D.** Let's use row annotations to "explore" and visually confirm their behavior. We can add annotations that show the log-fold changes that we used for filtering.

**E.** Finally, let's combine all the elements into a single heatmap.

::: {.callout-tip collapse="true"}

### Solution
```{r}
# A: Prepare the data
mat <- as.matrix(data$counts)
mat_scaled <- t(scale(t(mat))) # Z-score scaling
gene_vars <- apply(mat_scaled, 1, var, na.rm = TRUE)
top_genes_var <- names(sort(gene_vars, decreasing = TRUE)[1:50])
mat_top_var <- mat_scaled[top_genes_var, ]

# B: Create a column annotation with custom colors
col_anno <- HeatmapAnnotation(
  compound = data$anno$compound,
  time = data$anno$time,
  col = list(
    compound = c("none" = "steelblue", "sucrose" = "tomato"),
    time = c("3 hour" = "lightgrey", "24 hour" = "darkgrey")
  )
)

# C: Extract genes matching the profile
profile_genes <- data$diff %>%
  filter(
    sucrose_3h_lfc > 1 & sucrose_3h_pval < 0.05 |
      sucrose_24h_lfc < -1 & sucrose_24h_pval < 0.05
  ) %>%
  pull(Gene_ID)

mat_profile <- mat_scaled[profile_genes, ]

# D: Create a row annotation
# First, get the LFC values for our profile genes
rownames(data$diff) <- data$diff$Gene_ID
lfc_data <- data$diff[profile_genes, c("sucrose_3h_lfc", "sucrose_24h_lfc")]

# The anno_barplot function requires all values in a row to have the same sign
# when given a matrix. To visualize both positive and negative LFCs for each
# gene, we create two separate barplot annotations, one for each time point.
row_anno <- rowAnnotation(
  "3h LFC" = anno_barplot(
    lfc_data[, 1],
    gp = gpar(fill = "tomato"),
    axis_param = list(at = c(-2, 0, 2))
  ),
  "24h LFC" = anno_barplot(
    lfc_data[, 2],
    gp = gpar(fill = "steelblue"),
    axis_param = list(at = c(-2, 0, 2))
  ),
  width = unit(4, "cm")
)

# E: Combine with the heatmap
Heatmap(mat_profile,
  name = "Z-Score",
  top_annotation = col_anno,
  show_row_names = FALSE,
  column_title = "Genes Up at 3h, Down at 24h"
) + row_anno
```
:::

---

## Exercise 2: Combining `ComplexHeatmap` and `ggplot`

In this exercise, we will combine our `ComplexHeatmap` with a `ggplot` scatter plot. This is a common task when you want to present a global view of the data (the scatter plot) along with a detailed view of a specific subset of the data (the heatmap).

**A.** First, create a `ggplot` scatter plot of `sucrose_3h_lfc` vs. `sucrose_24h_lfc` for all genes.

**B.** On the scatter plot, color the genes that you selected in Exercise 1 to highlight them.

**C.** Use the `grid` package to combine the `ggplot` and the `ComplexHeatmap` from Exercise 1.

::: {.callout-tip collapse="true"}

### Solution
This is a complex task. `ComplexHeatmap` and `ggplot2` use different graphics systems. We can't use `patchwork`. We must use the base `grid` system to arrange them.

```{r, fig.height=6, fig.width=10}
# A & B: Create the ggplot
p_scatter <- ggplot(data$diff, aes(x = sucrose_3h_lfc, y = sucrose_24h_lfc)) +
  geom_point(color = "grey", alpha = 0.5) +
  geom_point(
    data = data$diff[profile_genes, ],
    color = "red",
    size = 2
  ) +
  theme_light() +
  labs(title = "LFC 3h vs 24h")

# C: Combine the plots
# 1. Draw the heatmap and get its object
ht <- Heatmap(mat_profile,
  name = "Z-Score",
  top_annotation = col_anno,
  show_row_names = FALSE,
  column_title = "Genes Up at 3h, Down at 24h"
) + row_anno

ht_grob <- grid.grabExpr(draw(ht))

# 2. Clear the page and set up a layout
grid.newpage()
pushViewport(viewport(layout = grid.layout(nrow = 1, ncol = 2, widths = unit(c(1, 1.5), "null"))))

# 3. Draw the ggplot in the first column
pushViewport(viewport(layout.pos.col = 1))
print(p_scatter, newpage = FALSE)
popViewport()

# 4. Draw the heatmap in the second column
pushViewport(viewport(layout.pos.col = 2))
grid.draw(ht_grob)
popViewport()
popViewport()
```
:::

## Setting a specific theme for plotting
```{r}
data_cols <- brewer.pal(n = 6, name = "Paired")

hm_cols <- list(
  compound = c(
    none = data_cols[1],
    sucrose = data_cols[2]
  ),
  dose = c(
    none = data_cols[3],
    "15 millimolar" = data_cols[4]
  ),
  time = c(
    "24 hour" = data_cols[6],
    "3 hour" = data_cols[5]
  )
)

# Theme ----
th <- theme_bw(base_family = "Helvetica") +
  theme(
    plot.title = element_text(
      colour = "black",
      angle = 0,
      size = 10,
      face = "bold",
      vjust = 1
    ),
    plot.subtitle = element_text(
      colour = "black",
      angle = 0,
      size = 10,
      face = "plain",
      vjust = 1
    ),
    plot.caption = element_text(
      colour = "black",
      angle = 0,
      size = 10,
      face = "plain",
      vjust = 1,
      hjust = 0.5
    ),
    axis.text.x = element_text(
      colour = "black",
      angle = 90,
      size = 9
    ),
    axis.text.y = element_text(
      colour = "black",
      angle = 0,
      size = 9
    ),
    axis.title.x = element_text(
      colour = "black",
      size = 10,
      face = "plain"
    ),
    axis.title.y = element_text(
      colour = "black",
      size = 10,
      face = "plain"
    ),
    legend.text = element_text(
      colour = "black",
      size = 9
    ),
    legend.title = element_text(
      size = 10,
      face = "plain",
      colour = "black"
    ),
    strip.background = element_rect(
      fill = NA,
      colour = "black"
    ),
    strip.text.x = element_text(
      face = "plain",
      colour = "black",
      size = 10
    ),
    strip.text.y = element_text(
      face = "plain",
      colour = "black",
      size = 10
    )
  ) +
  theme(
    panel.background = element_rect(
      fill = "white"
    ),
    panel.border = element_rect(
      colour = "black",
      fill = NA
    ),
    strip.background = element_rect(
      color = "black"
    )
  )
```

## Exercise 3: Including external images and `gganatogram` plots

In this exercise, we will include an image from the web and create a `gganatogram` plot.

**A.** Include the image from the web.

**B.** Create a `gganatogram` plot of mouse respiratory system.

```{r}
# A: Include image from the web
img <- "https://mdpi-res.com/ijms/ijms-22-04704/article_deploy/html/images/ijms-22-04704-g004.png"
img_draw <- ggdraw() + draw_image(img)

img_draw

# B: Create gganatogram plot
mm_res <- mmMale_key %>%
  dplyr::filter(type == "respiratory") %>%
  gganatogram(outline = T, fillOutline = "#a6bddb", organism = "mouse", sex = "male", fill = "colour") + 
  theme_bw() + 
  coord_fixed()

mm_res
```

## Exercise 4: Advanced setting of height and width of indivisual plots and combining all plots

In this exercise, we will combine all the plots we have created so far into a single panel.

**A.** Set the height and width of the scatter plot.

**B.** Combine the scatter plot, heatmap, web image, and gganatogram plot into a single panel.

::: {.callout-tip collapse="true"}

### Solution
```{r}
# A: Set height and width of the scatter plot
p1 <- gridExtra::grid.arrange(
  egg::set_panel_size(
    p = p_scatter + th, width = unit(4, "cm"), height = unit(4, "cm")
    )
  )

# B: Combine all plots
gridExtra::grid.arrange(p1, img_draw, ht_grob, mm_res, ncol = 2, )
```
:::

## Session information
::: {.callout-tip collapse="true"}
```{r}
sessionInfo()
```
:::